Parsing Postfix log files

### 2008-07-26 doc/logparser-results.tex
Lots of improvements to the efficiency section - I'm happy with most of it now. I improved the graphs too, with better captions and labels.

XXX CHANGE normal TO optimal IN ORDERING STUFF, INCLUDING IN GRAPH CAPTIONS
AND AXES LABELS\@.

@@

XXX THIS SECTION NEEDS TO BE REWRITTEN\@.

Parsing efficiency is an obvious concern when the parser routinely needs to
parse very large log files.  The server which generated the log files used
in testing this parser accepts approximately 10,000 mails for 700 users a
day, and on average it produces 75MB log files containing 300,000 log lines
--- large scale mail servers would have much larger log files on a daily
basis.

\showgraph{build/plot-mails-received}{Mails received per day}{Mails
received per day}

As expected there are far more mails received on weekdays than at weekends.
Note that this graph and the table below show the number of mails received
by \SMTP{} only; in particular the mail loops noticeable in other graphs
do not contribute to these figures.

\showtable{build/mails-received-include-for-graph}{Number of mails received
per day: statistics}{Number of mails received per day: statistics}

When generating the timing data used in this section, \numberOFlogFILES{}
log files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
parsed 10 times, the results of the first first run discarded, and the
execution time for the remaining 9 runs averaged.  The results from the
first run are discarded for two reasons:

    \item The execution time will be higher because the log file must be
        read from disk, whereas for subsequent runs the log file will be
        cached in memory by the operating system.

    \item The execution time will also be higher because the rule ordering
        will be sub-optimal compared to subsequent runs.

Saving results to the database was disabled for the test runs, as that
dominates the run time of the program, and the tests are aimed at measuring
the speed of the parser rather than the speed of the database and the disks
the database is stored on.

\newpage

@@

An important property of a parser is how execution time scales relative to
input size: does it scale linearly, polynomially, or exponentially?
\Graphref{execution time vs file size vs number of lines graph} shows the
execution time in seconds, file size in MB, and tens of thousands of log
lines per log file.  All three lines run roughly in parallel, giving the
impression that the algorithm scales linearly with input size.  This
impression is borne out by \graphref{execution time vs file size vs number
lines factor} which plots both the ratio of file size vs execution time,
and the ratio of number of log lines vs execution time (higher is better in
both cases); \tableref{execution time vs file size vs number lines factor
table} shows the ratios for different groups of log files.  As the
reader can see the ratios are quite tightly banded, showing that the
algorithm scales linearly: the much larger log files between points 60 and
70 on the X axis in \graphref{execution time vs file size vs number of
lines graph} actually cause the ratio to increase (i.e.\ improve), rather
than decrease.  The strange behaviour where larger log files are parsed
more efficiently is explained fully in \sectionref{Explaining the peaks in
log file size}.

\showgraph{build/plot-normal-filesize-numlines}{Execution time vs file
size vs number of lines}{execution time vs file size vs number of lines
graph}

The Y axis in \graphref{execution time vs file size vs number of lines
graph} represents the following:

    \item The time required, in seconds, to parse the log file.

    \item The size of the log file, in megabytes.

    \item The number of lines in the log file, divided by 10000.

\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file
size and number of lines to execution time}{execution time vs file size vs
number lines factor}

\showtable{build/stats-normal-filesize-line-count-include}{Ratio of file
size \& number of lines to execution time: statistics}{execution time vs
file size vs number lines factor table}

@@

\Graphref{execution time vs file size vs number of lines graph} shows log
file size varying between 30--100MB, except for log files 22 (200MB) and
62--68 (400--900MB).  These large log files were caused by a user
unintentionally creating a mail forwarding loop, resulting in a huge number
of mail deliveries and up to an order of magnitude more log lines than is
usual.  The user specified his mail forwarding as: \newline
\tab{}\texttt{$\backslash$username, username@domain} \newline This
instructs Postfix to deliver the mail to the local user, and also forward
it to the remote address; this is generally not a problem except that the
remote address in this case is the address the mail was originally sent to,
creating an infinite loop.  To prevent this happening Postfix examines the
Delivered-To header in the mail, and if the mail has already been delivered
to the current address it is bounced back to the sender with the error
message \texttt{mail forwarding loop for username@domain}.  Ordinarily this
works well, but unfortunately in this case the user noticed they had not
received any mail in a while and opted to send a test mail to themselves,
causing a loop not caught by Postfix:

@@

At the time of writing there are \numberOFrules{} different rules: the top
10\% match the vast majority of log lines, with the remaining log lines
split across the other 90\% of the rules (as shown in \graphref{rule hits
graph}).  Assuming that the distribution of log lines is reasonably steady
over time, parser efficiency should benefit from trying more frequently
matching rules before those which match less frequently.  To test this
hypothesis three full test runs were performed with different rule
orderings:

@@

    \item [shuffle] This is intended to represent a randomly ordered rule
        set.  The rules will be shuffled once before use and will retain
        that ordering for the entirety of the log file.  Note that the
        ordering will change every time the parser is executed, so 10
        different orderings will be generated for each log file in the test
        run.

@@

\Graphref{Execution time relative to
shuffled ordering graph} shows the execution time of normal and reverse orderings
relative to shuffled ordering; the mean increases for different groupings
of log files are given in \tableref{Execution time relative to shuffled
ordering table}.  Overall this optimisation provides a modest but worthwhile
performance increase of approximately 10\%, for a small investment in time
and programming.

\showgraph{build/plot-normal-and-reverse-relative-to-shuffle}{Execution
time of normal and reverse orderings relative to shuffled
ordering}{Execution time relative to shuffled ordering graph}

\showtable{build/stats-normal-shuffle-reverse-include}{Execution time
relative to shuffled ordering}{Execution time relative to shuffled ordering
table}

XXX IMPROVE THE REMAINDER OF THIS SECTION\@; CLEARLY SEPARATE THE
CHARACTERISTICS AND CLARIFY THEM\@.

Different rule orderings have less effect when parsing log files 22 and
62--68, due to the different distribution of log lines in those log files
(see \sectionref{Explaining the peaks in log file size}).  The vast
majority of log lines when a mail loop occurs are from Postfix components
which have a small number of rules associated with them, whereas in general
\daemon{smtpd} adds the majority of log lines, and also has the highest
number of rules.  \daemon{smtpd} log lines are distributed across rules
much more evenly than the log lines of \daemon{qmgr}, \daemon{local},
\daemon{cleanup} or \daemon{pickup}, so the average number of rules
required to parse a \daemon{smtpd} log line is much higher than the average
number required to parse other log lines.

These two characteristics combine to reduce the average number of rules
required to parse a log line when there is a mail loop.  The ratio of
number of lines versus execution time in \graphref{execution time vs file
size vs number lines factor} shows a noticeable improvement at log files 22
and 62--68, where a higher ratio means more log lines parsed per unit of
execution time.

When the rule ordering is reversed the majority of log lines generated by a
mail loop will be parsed with very few rules, whereas without a mail loop
the majority of log lines require a large number of rules; this leads to a
noticeable drop in the average time required to parse a log line, as shown
in \graphref{Execution time relative to shuffled ordering graph}.  The
number of rules which need to be consulted when the ordering is shuffled
varies between the optimal and the pessimal, and the performance varies
proportionally.

@@

Before Perl attempts to match a \regex{} against a piece of text, the
original \regex{} is compiled into an internal representation and optimised
to improve the speed of matching, but this compilation and optimisation
takes CPU time; far more CPU time, in fact, than the actual matching takes
in many cases.  Perl automatically caches static \regexes{} at compile
time, but \regexes{} loaded dynamically while the program need to be
explicitly compiled and cached.

The following graphs show the impact that not caching compiled \regexes{}
has on parser performance: on typical log files the execution time when not
caching compiled \regexes{} is 500--600\% of the execution time when
caching; reversing the perspective shows that cached execution time is
merely 17--20\% of non-cached execution time.  Caching compiled \regexes{}
is probably the single most effective optimisation possible in the parser's
implementation; given that it only required sixteen extra lines of
reasonably simple code,\footnote{The sixteen lines of code breaks down as
follows: three lines to add caching, eight lines of error checking and
reporting, three lines to optionally disable caching for debugging and
performance measurement, and two lines of comments.} the investment in time
was minimal.

\showgraph{build/plot-cached-discarded}{Regex: cached vs
discarded}{normal regex vs discard regex}

\showgraph{build/plot-cached-discarded-factor}{Regex caching: percentage
execution time increase}{normal regex vs discarded regex factor}

Two large dips can be seen in \graphref{normal regex vs discarded
regex factor} at log files 22 and 62--68, corresponding to the spikes in
log file size in \graphref{execution time vs file size vs number
of lines graph} and \graphref{normal regex vs discard regex}.  The
reason for the anomalous log files has already been explained in
\sectionref{Explaining the peaks in log file size}.

The distribution of log lines across rules when there is a mail loop is
much different to the norm and the average number of rules consulted per
log line is much lower; this results in far fewer \regex{} compilations per
line than when there is not a mail loop, and a correspondingly decreased
execution time.  The unexpected outcome is that caching \regexes{} is
proportionally less important when the log file contents were created by a
mail loop.  The increases in execution time when not caching \regexes{} for
different combinations of log files are summarised in the table below:

XXX THIS IS AN AWFUL CAPTION\@.

\showtable{build/stats-cached-discarded-include-for-graph}
{Regex caching/discarding with different groups of log files}
{Regex caching/discarding with different groups of log files}

@@

>>>>

XXX CHANGE normal TO optimal IN ORDERING STUFF, INCLUDING IN GRAPH CAPTIONS
AND AXIS LABELS\@.

@@

Parsing efficiency is an obvious concern when the parser routinely needs to
parse very large log files.  The server which generated the log files used
in testing this parser accepts approximately 10,000 mails for 700 users a
day; as expected there are far more mails received on weekdays than at
weekends.   Median log file size is 50MB, containing 285,000 log lines ---
large scale mail servers would have much larger log files.  Note that
\graphref{Mails received per day} and \tableref{Number of mails received
per day: statistics} show the number of mails received by \SMTP{} only; in
particular the mail loops noticeable in other graphs do not contribute to
these figures.

\showgraph{build/plot-mails-received}{Number of mails received per
day}{Mails received per day}

\showtable{build/mails-received-include-for-graph}{Number of mails received
per day}{Number of mails received per day: statistics}

When generating the timing data used in this section, \numberOFlogFILES{}
log files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
parsed 10 times, the results from the first first run discarded, and the
execution time for the remaining 9 runs averaged.  The results from the
first run were discarded because because the rule ordering will be
sub-optimal compared to subsequent runs --- XXX MAYBE I SHOULDN'T DISCARD
THE FIRST RESULT\@?  Saving results to the database was disabled for the
test runs, as that dominates the run time of the program, and the tests are
aimed at measuring the speed of the parser rather than the speed of the
database and the disks the database is stored on.

@@

An important property of a parser is how execution time scales relative to
input size: does it scale linearly, polynomially, or exponentially?
\Graphref{parsing time vs file size vs number of lines graph} shows the
parsing time in seconds, file size in MB, and tens of thousands of log
lines for each of the \numberOFlogFILES{} log files.  All three lines run
roughly in parallel, giving the impression that the algorithm scales
linearly with input size.  This impression is borne out by
\graphref{parsing time vs file size vs number lines factor} which plots
both the ratio of file size vs parsing time, and the ratio of number of log
lines vs parsing time (higher is better in both cases); \tableref{parsing
time vs file size vs number lines factor table} shows the same ratios for
different groups of log files.  As the reader can see the ratios are quite
tightly banded, providing empirical evidence that the algorithm scales
linearly.  The ratio increases (i.e.\ improves) for log files 22 and 62--68
despite their larger than average size (shown in \graphref{parsing time vs
file size vs number of lines graph}); this strange behaviour where larger
log files are parsed more efficiently than smaller is explained fully in
\sectionref{Explaining the peaks in log file size}.

\showtable{build/stats-normal-filesize-line-count-include}{Ratio of file
size and number of lines to parsing time}{parsing time vs file size vs
number lines factor table}

\showgraph{build/plot-normal-filesize-numlines}{Parsing time, file size,
and number of lines}{parsing time vs file size vs number of lines graph}

\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file size
and number of lines to parsing time}{parsing time vs file size vs number
lines factor}

@@

\Graphref{parsing time vs file size vs number of lines graph} shows log
file size varying between 30--100MB, except for log files 22 (200MB) and
62--68 (400--900MB).  These large log files were caused by a user
unintentionally creating a mail forwarding loop, resulting in a huge number
of mail deliveries and up to an order of magnitude more log lines than is
usual.  The user specified his mail forwarding as: \newline
\tab{}\texttt{$\backslash$username, username@domain} \newline This
instructs Postfix to deliver the mail to the local user, and also forward a
copy to the remote address; this is generally not a problem except that the
remote address in this case is the address the mail was originally sent to,
creating an infinite loop.  To prevent this happening Postfix examines the
Delivered-To header in the mail, and if the mail has already been delivered
to the current address it is bounced back to the sender with the error
message \newline{} \tab{} \texttt{mail forwarding loop for
username@domain}\newline{}  Ordinarily this
works well, but unfortunately in this case the user noticed they had not
received any mail in a while and opted to send a test mail to themselves,
causing a loop not caught by Postfix:

@@

At the time of writing there are \numberOFrules{} different rules: the top
10\% match the vast majority of log lines, with the remaining log lines
split across the other 90\% of the rules (as shown in \graphref{rule hits
graph}).  Assuming that the distribution of log lines is reasonably
consistent over time, parser efficiency should benefit from trying more
frequently matching rules before those which match less frequently.  To
test this hypothesis three full test runs were performed with different
rule orderings:

@@

    \item [shuffle] This is intended to represent a randomly ordered rule
        set.  The rules will be shuffled once before use and will retain
        that ordering until the parser exits.  Note that the ordering will
        change every time the parser is executed, so 10 different rule
        orderings will be generated for each log file in the test run.

@@

\Graphref{Parsing time relative to shuffled ordering graph} shows the
execution time of normal and reverse orderings relative to shuffled
ordering; the mean relative execution times for different groupings of log
files are given in \tableref{Parsing time relative to shuffled ordering
table}.  Overall this optimisation provides a modest but worthwhile
performance increase of approximately 10\%, for a small investment in time
and programming.

\showgraph{build/plot-normal-and-reverse-relative-to-shuffle}{Parsing time
relative to shuffled ordering}{Parsing time relative to shuffled ordering
graph}

\showtable{build/stats-normal-shuffle-reverse-include}{Parsing time
relative to shuffled ordering}{Parsing time relative to shuffled ordering
table}

Differences in rule ordering have less effect on parsing time when parsing
log files 22 and 62--68, due to the different distribution of log lines in
those log files (see \sectionref{Explaining the peaks in log file size}).
When a mail loop occurs the vast majority of log lines are from Postfix
components which have a small number of rules associated with them, whereas
in general the majority of log lines are from Postfix components which have
a high number of associated rules; this results in the average number of
rules required to parse a log line from log files 22 or 62--68 being
considerably lower than usual.  The ratio of number of lines versus parsing
time in \graphref{parsing time vs file size vs number lines factor} shows a
noticeable improvement at log files 22 and 62--68, where a higher ratio
means more log lines parsed per unit of time; similarly \graphref{Parsing
time relative to shuffled ordering graph} shows much less difference
between the orderings at log files 22 and 62--68.

@@

Before the perl interpreter attempts to match a \regex{} against a piece of
text, the \regex{} is compiled into an internal representation and
optimised to improve the speed of matching.  This compilation and
optimisation takes CPU time ---in many cases it takes far more CPU time
than the actual matching.  If the interpreter is certain that a \regex{}
will not change it will automatically cache the compilation results for
later use.   The results of compiling a dynamically generated \regex{} can
be cached and used in preference to the original \regex{}, but it is the
responsibility of the programmer to do this.  The Postfix parser loads the
rules it uses from a database, and thus each rule's \regex{} is compiled
and the result cached to reduce parsing time.

\Graphref{Increase in parsing time when not caching compiled regexes graph}
shows the impact that not caching compiled \regexes{} has on parser
performance: on typical log files the parsing time when not caching
compiled \regexes{} is 400--600\% of the parsing time when caching.
Caching compiled \regexes{} is probably the single most effective
optimisation possible in the parser's implementation, and was quite simple
to implement.

\showgraph{build/plot-cached-discarded-factor}{Increase in parsing time
when not caching compiled regexes}{Increase in parsing time when not
caching compiled regexes graph}

Two large dips can be seen in \graphref{Increase in parsing time when not
caching compiled regexes graph} at log files 22 and 62--68, corresponding
to the spikes in log file size in \graphref{parsing time vs file size vs
number of lines graph}.  The distribution of log lines across rules when
there is a mail loop is much different to the norm, and the average number
of rules consulted per log line is much lower: this results in far fewer
\regex{} compilations per line when there is a mail loop, and a
corresponding decrease in the average parsing time for a single log line.
The outcome is that caching compiled \regexes{} is proportionally less
important when the log file contents were created by a mail loop.  The
increases in parsing time when not caching compiled \regexes{} for
different groups of log files are summarised in \tableref{Increase in
parsing time when not caching compiled regexes table}.

\showtable{build/stats-cached-discarded-include-for-graph} {Increase in
parsing time when not caching compiled regexes} {Increase in parsing time
when not caching compiled regexes table}

@@

XXX FIGURE OUT WHERE TO PUT THESE GRAPHS\@; DO I NEED ALL THREE\@?
